---
title: "ColBioProject_ZIMMERMAN_FRENZE_STONEFIELD_LANG_URBANOWICZ_MORALES"
format: html
editor: visual
---

## Load Packages

```{r}
# Everyone as needed:
library(tidyverse)
```

## Load Data

```{r}
# Ian
# Pull data from https://github.com/Stoney044/Ebird_IPEDS

EB_url <- "https://raw.githubusercontent.com/Stoney044/Ebird_IPEDS/main/EBbird.csv"
IP_url <- "https://raw.githubusercontent.com/Stoney044/Ebird_IPEDS/main/IPEDS.csv"

# Load the data
Ebird <- read_csv(EB_url)
IPEDS <- read_csv(IP_url)

# Clean Metadata
# Ian
metadata <- metadata |>
  select(varTitle, varName) 

IPEDS_columns <- IPEDS |> 
  colnames()

# Filter the metadata only to include the columns in the IPEDS data
metadata <- metadata |>
  filter(varName %in% IPEDS_columns)

# export metadata as csv file 
# write_csv(metadata, "Data/Clean_IPEDS_metadata.csv")
```

## Bird Species Richness per Campus (2)

### Richness Per Sampling Event

```{r}
# Ingrid, Lee, Elizabeth, Jake, Ian

species_richness <- Ebird |> 
  group_by(UNITID, `SAMPLING EVENT IDENTIFIER`) |> 
  mutate(species_count = n_distinct(`SCIENTIFIC NAME`))

```

### Average Sampling Event Richness per UNITID

```{r}
# Ingrid, Lee, Elizabeth, Jake, Ian

mean_species_richness <- species_richness |> 
  group_by(UNITID) |> 
  summarise(mean_species_richness = mean(species_count, na.rm = TRUE))

```

## Join Species Richness to ipeds data (3)

```{r}
# Elizabeth

merged_data <- IPEDS |> 
  left_join(mean_species_richness)

```

## Explore bird species richness and impervious surface (4)

### Hypothesis (4a)

There will be less richness with more impervious surface because these surfaces cannot provide the resources birds need to survive.

### Relationship Across Colleges (4b)

```{r}
# Lee

merged_data |>
  group_by(UNITID) |> 
  ggplot(aes(x = mean_impervious, y = mean_species_richness)) +
  geom_smooth(se = FALSE)
  
merged_data |>
  group_by(UNITID) |> 
  ggplot(aes(x = log(mean_impervious), y = mean_species_richness)) +
  geom_smooth(se = FALSE)  

merged_data |>
  group_by(UNITID) |> 
  ggplot(aes(x = mean_impervious, y = log(mean_species_richness))) +
  geom_smooth(se = FALSE)

merged_data |>
  group_by(UNITID) |> 
  ggplot(aes(x = log(mean_impervious), y = log(mean_species_richness))) +
  geom_smooth(se = FALSE)

```

### Mechanism (4c)

Mostly stable then dips, implies that there is a threshold of impervious surface that is okay before it causes problems.

### By Taxonomic Order (4d)


```{r}
# Lee

Ebird |> 
  group_by(UNITID, ORDER) |> 
  mutate(species_count = n_distinct(`SCIENTIFIC NAME`)) |> 
  group_by(UNITID, ORDER) |> 
  summarise(mean_species_richness = mean(species_count, na.rm = TRUE)) |> 
  right_join(IPEDS) |> 
  ggplot(aes(x = mean_impervious, y = mean_species_richness)) +
  geom_smooth(se = FALSE) +
  facet_wrap(~ORDER, scales = "free_y")

```

## Team Member Individual Explorations

### Zimmerman

### Frenze

### Stonefield

### Lang

species_richness <- Ebird |>
  group_by(UNITID, `SAMPLING EVENT IDENTIFIER`) |>
  mutate(species_count = n_distinct(`SCIENTIFIC NAME`))
mean_species_richness <- species_richness |>
  group_by(UNITID) |>
  summarise(mean_species_richness = mean(species_count, na.rm = TRUE))
merged_data <- IPEDS |>
  left_join(mean_species_richness)

library(tidyverse)
library(maps)
library(viridis)
library(ggthemes)
# Load the US state map data
us <- map_data("state")
lat_long <- merged_data |>
  select(mean_species_richness, UNITID, IPEDS_LATITUDE, IPEDS_LONGITUDE)
# Create the plot
ggplot(lat_long) +
  geom_polygon(data = us, aes(x = long, y = lat, group = group), color = 'gray1', fill = NA, alpha = .35) +
  geom_point(aes(x = IPEDS_LONGITUDE, y = IPEDS_LATITUDE, color = mean_species_richness), size = 1, alpha = 0.5) +
  scale_color_viridis(option = "C", direction = -1, begin = 0.3, end = 0.9)+
xlim(-130, -65) +
  ylim(24, 50) +
  labs(
    title = "Mean Species Richness Across US Colleges",
    subtitle = "Bird species richness by college campus location",
    color = "Species Richness",
    caption = "Data source: eBird & IPEDS"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    plot.subtitle = element_text(size = 16),
    plot.caption = element_text(size = 8),
    axis.text = element_blank(), 
    axis.ticks = element_blank(), 
    axis.title = element_blank() 
  )


# 4b: Hypothesis is that larger cities have an increase bird
# species richness. Larger cities may have more resources and
# varied habitats, such as parks and green spaces, which can
# support a greater diversity of bird species with more habitats
# and more variety. 

# 4c: when accounting for impervious surface, it could change these 
# results as the urbanization of college campuses could also have 
# a negative effect on bird species richness. Examining taxonomic 
# order can also affect the species richness depending on if 
# certain groups are more or less affected by the urban college
# campuses

### Urbanowicz

### Morales
