---
title: "ColBioProject_ZIMMERMAN_FRENZE_STONEFIELD_LANG_URBANOWICZ_MORALES"
format: html
editor: visual
---

## Load Packages

```{r}
# Everyone as needed:
library(tidyverse)
```

## Load Data

```{r}
# Ian
# Pull data from https://github.com/Stoney044/Ebird_IPEDS

EB_url <- "https://raw.githubusercontent.com/Stoney044/Ebird_IPEDS/main/EBbird.csv"
IP_url <- "https://raw.githubusercontent.com/Stoney044/Ebird_IPEDS/main/IPEDS.csv"

# Load the data
Ebird <- read_csv(EB_url)
IPEDS <- read_csv(IP_url)

# Clean Metadata
# Ian
metadata <- metadata |>
  select(varTitle, varName) 

IPEDS_columns <- IPEDS |> 
  colnames()

# Filter the metadata only to include the columns in the IPEDS data
metadata <- metadata |>
  filter(varName %in% IPEDS_columns)

# export metadata as csv file 
# write_csv(metadata, "Data/Clean_IPEDS_metadata.csv")
```

## Bird Species Richness per Campus (2)

### Richness Per Sampling Event

```{r}
# Ingrid, Lee, Elizabeth, Jake, Ian

species_richness <- Ebird |> 
  group_by(UNITID, `SAMPLING EVENT IDENTIFIER`) |> 
  mutate(species_count = n_distinct(`SCIENTIFIC NAME`))

```

### Average Sampling Event Richness per UNITID

```{r}
# Ingrid, Lee, Elizabeth, Jake, Ian

mean_species_richness <- species_richness |> 
  group_by(UNITID) |> 
  summarise(mean_species_richness = mean(species_count, na.rm = TRUE))

```

## Join Species Richness to ipeds data (3)

```{r}
# Elizabeth

merged_data <- IPEDS |> 
  left_join(mean_species_richness)

```

## Explore bird species richness and impervious surface (4)

### Hypothesis (4a)

There will be less richness with more impervious surface because these surfaces cannot provide the resources birds need to survive.

### Relationship Across Colleges (4b)

```{r}
# Lee




merged_data |>
  group_by(UNITID) |> 
  ggplot(aes(x = mean_impervious, y = mean_species_richness)) +
  geom_smooth(se = FALSE) + labs(x = "Mean Impervious Surface", y = "Mean Species Richness")
  
merged_data |>
  group_by(UNITID) |> 
  ggplot(aes(x = log(mean_impervious), y = mean_species_richness)) +
  geom_smooth(se = FALSE) + labs(x = "Log(Mean Impervious Surface)", y = "Mean Species Richness")  

merged_data |>
  group_by(UNITID) |> 
  ggplot(aes(x = mean_impervious, y = log(mean_species_richness))) +
  geom_smooth(se = FALSE) + labs(x = "Mean Impervious Surface", y = "Log(Mean Species Richness)")

merged_data |>
  group_by(UNITID) |> 
  ggplot(aes(x = log(mean_impervious), y = log(mean_species_richness))) +
  geom_smooth(se = FALSE) + labs(x = "Log(Mean Impervious Surface", y = "Log(Mean Species Richness)")

```

### Mechanism (4c)

Mostly stable then dips, implies that there is a threshold of impervious surface that is okay before it causes problems.

### By Taxonomic Order (4d)


```{r}
# Lee

Ebird |> 
  group_by(UNITID, ORDER) |> 
  mutate(species_count = n_distinct(`SCIENTIFIC NAME`)) |> 
  group_by(UNITID, ORDER) |> 
  summarise(mean_species_richness = mean(species_count, na.rm = TRUE)) |> 
  right_join(IPEDS) |> 
  ggplot(aes(x = mean_impervious, y = mean_species_richness)) +
  geom_smooth(se = FALSE) +
  facet_wrap(~ORDER, scales = "free_y")

```

## Team Member Individual Explorations

### Zimmerman

### Frenze
```{r}
merged_data_NF<-merged_data|>select(
  INSTNM, INSTSIZE, ENRTOT, mean_impervious, mean_species_richness)|>
  filter(!is.na(mean_species_richness))|>
  filter(!is.na(ENRTOT))|>
  mutate(INSTSIZE = factor(INSTSIZE))

Enrollment_range<-c("Under 1,000", "1,000-5,000", "5,000-10,000", "10,000-20,000", "20,000+")

plot_1NF<-ggplot(data = merged_data_NF) +
  geom_boxplot(aes(x = INSTSIZE, y = mean_species_richness, color = INSTSIZE)) +
  labs(x = "Institutation Size", y = "Mean Species Richness", color = "Number of Students Enrolled") + 
  theme(legend.position = "bottom") +
  scale_color_discrete(labels = Enrollment_range)  

species_richness_per_enrollment<-merged_data_NF|>mutate(
  Species_richness_per_student = mean_species_richness/ENRTOT,
  .keep = "all"
)

plot_2NF<-ggplot(data = species_richness_per_enrollment) +
  geom_boxplot(aes(x = INSTSIZE, y = Species_richness_per_student, color = INSTSIZE)) +
  labs(x = "Institutation Size", y = "Mean Species Richness per Student", color = "Number of Students Enrolled") + 
  theme(legend.position = "none") +
  scale_color_discrete(labels = Enrollment_range) 
print(plot_1NF + plot_2NF)
```

There is no apparent correlation between the average impervious surface or institution size. This data set suggest there is little correlation between institution size and species richness. \# However it should be noted that seeing that even the smaller institution sizes have equitable species richness the larger ones which would logically require a larger campuses on average. This could indicate that smaller institutions may have higher species richness per square mile and with additional data pertaining to the each individual campuses area further analysis can be done.
### Stonefield

#### Mean Species Richness based on Institution's Land Grant Status: 

Hypothesis: Land grant universities were established due to a grant program from the federal government to teach agriculture, military tactics, and the mechanic arts as well as classical studies so members of the working classes could obtain a liberal, practical education. Due to this, Land grant universities tend to be in more rural areas, therefor should have a higher mean species richness.

```{r}
#| label: Land Grant Status
#| warning: false
land_grant_SR <- merged_data |> 
  select(UNITID, INSTNM, mean_species_richness, LANDGRNT)|>
  filter(is.na(mean_species_richness) == FALSE) |>
  mutate(LANDGRNT = ifelse(LANDGRNT == 1, "Land Grant", "Not Land Grant"))

land_grant_SR |>
  ggplot(aes(x = LANDGRNT, y = mean_species_richness)) +
  geom_boxplot(aes(fill = LANDGRNT)) +
  labs(title = "Mean Species Richness of Land Grant Universities",
       x = "Land Grant Status",
       y = "Mean Species Richness") +
  theme_minimal() +
  scale_fill_manual(values = c("Land Grant" = "green", "Not Land Grant" = "skyblue2"))

```

Impervious surfaces may be more prevalent on non-land-grant institutions so this may be a factor in these results. Taxonomic order breakdown is shown below.

```{r}
#| label: by taxonomic order
#| warning: false
# Faceted by Order

mean_species_richness <- species_richness |> 
  group_by(UNITID, ORDER) |> 
  summarise(mean_species_richness = mean(species_count, na.rm = TRUE))

merged_data <- IPEDS |> 
  left_join(mean_species_richness)

land_grant_SR <- merged_data |> 
  select(UNITID, INSTNM, mean_species_richness, LANDGRNT, ORDER)|>
  filter(is.na(mean_species_richness) == FALSE) |>
  mutate(LANDGRNT = ifelse(LANDGRNT == 1, "Land Grant", "Not")) |> 
  filter(!is.na(ORDER))

land_grant_SR |>
  group_by(ORDER) |>
  ggplot(aes(x = LANDGRNT, y = mean_species_richness)) +
  geom_boxplot(aes(fill = LANDGRNT)) +
  labs(title = "Mean Species Richness of Land Grant Universities",
       x = "Land Grant Status",
       y = "Mean Species Richness") +
  theme_minimal() +
  scale_fill_manual(values = c("Land Grant" = "green", "Not" = "skyblue2")) +
  facet_wrap(~ORDER)
```

### Lang
```{r}
species_richness <- Ebird |>
  group_by(UNITID, `SAMPLING EVENT IDENTIFIER`) |>
  mutate(species_count = n_distinct(`SCIENTIFIC NAME`))
mean_species_richness <- species_richness |>
  group_by(UNITID) |>
  summarise(mean_species_richness = mean(species_count, na.rm = TRUE))
merged_data <- IPEDS |>
  left_join(mean_species_richness)

library(tidyverse)
library(maps)
library(viridis)
library(ggthemes)
# Load the US state map data
us <- map_data("state")
lat_long <- merged_data |>
  select(mean_species_richness, UNITID, IPEDS_LATITUDE, IPEDS_LONGITUDE)
# Create the plot
ggplot(lat_long) +
  geom_polygon(data = us, aes(x = long, y = lat, group = group), color = 'gray1', fill = NA, alpha = .35) +
  geom_point(aes(x = IPEDS_LONGITUDE, y = IPEDS_LATITUDE, color = mean_species_richness), size = 1, alpha = 0.5) +
  scale_color_viridis(option = "C", direction = -1, begin = 0.3, end = 0.9)+
xlim(-130, -65) +
  ylim(24, 50) +
  labs(
    title = "Mean Species Richness Across US Colleges",
    subtitle = "Bird species richness by college campus location",
    color = "Species Richness",
    caption = "Data source: eBird & IPEDS"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    plot.subtitle = element_text(size = 16),
    plot.caption = element_text(size = 8),
    axis.text = element_blank(), 
    axis.ticks = element_blank(), 
    axis.title = element_blank() 
  )
```

4b: Hypothesis is that larger cities have an increase bird species richness. Larger cities may have more resources and varied habitats, such as parks and green spaces, which can support a greater diversity of bird species with more habitats and more variety. 

4c: when accounting for impervious surface, it could change these results as the urbanization of college campuses could also have a negative effect on bird species richness. Examining taxonomic order can also affect the species richness depending on if certain groups are more or less affected by the urban college campuses

### Urbanowicz

### Morales
```{r}
ICLEVEL <- merged_data |>
select(UNITID, INSTNM, mean_species_richness, ICLEVEL) |>
  mutate(ICLEVEL = case_when(
    ICLEVEL == 1 ~ "4-or-more-year",
    ICLEVEL == 2 ~ "2-year"
  ))
ICLEVEL |>
  ggplot(aes(x =ICLEVEL, y = mean_species_richness)) +
  geom_boxplot(aes(fill = ICLEVEL)) +
  labs(title = "Mean Species Richness of 4-or-more years and 2-year Institution",
  x = "Institution Level",
  y = "Mean Species richness") +
  theme_minimal() +
  scale_fill_manual(values = c("4-or-more-year" = "purple", "2-year" = "pink"))
```  
Hypothesis: Exploring the relationship between institution level and species richness.

The hypothesis that i want to test is that if there is relationship between the species richness and the institution  level. Since institutions often vary in size, location and influence they may have different footprints and may impact the local biodiversities from their surrounding environment. 
When we look at additional factors such as impervious surfaces and taxonomic order it can definitively change the result and might even provide a more clear understanding of the relationship between the species tichness and the institution level

